---
author: "Akshat Dwivedi"
date: "30-05-2016"
output: 
  pdf_document: 
    fig_caption: yes
    fig_width: 5
    highlight: monochrome
title: "Statistical Modelling Final Assignment"
html_document:
  highlight: monochrome
  keep_md: yes
  theme: journal
  toc: no
  toc_depth: 2
  toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, warning = FALSE)
#If you run into problems with cached output you can always clear the knitr cache by removing the folder named with a _cache suffix within your document’s directory.
```


# Part A




# Part B

## B.2: Simulation Study for Poisson Regression
This part involves conducting a simulation study that compares AIC, BIC and TIC for Poisson regression. Five variables, $x_{i,j}$, $j = 2,\ldots,6$, $i = 1,\ldots,n$ were sampled from Uniform(-1,1). The response variable $Y$ was sampled from Poisson($\mu_i$) with $$\mu_i = \text{exp}\left(\frac{1}{3} + \frac{2}{3}x_{i,2} + 1*x_{i,3} + \frac{4}{3}x_{i,4} + 0*x_{i,5} + 0*x_{i,6}\right)$$
The response vector is $n$ x $1$ and the design matrix is an $n$ x $6$ matrix with the first column containing the intercept (all 1's) and the rest of the columns containing the $x_j$'s. Furthermore, we construct models with all the possible combinations of the covariates (which for 5 covariates results in $2^5=32$ models) and keeping the intercept term in each of the models.

This simulation is run 1000 times for two sample sizes, $n=50$ and $n=400$. The aim is to see which models are selected and with what frequency by AIC, BIC and TIC. The results are presented below.









```{r ques_b, include=FALSE}
load(file = "/home/ad/Desktop/KUL Course Material/Statistical Modelling/R programming/ques_b_results.RData")

table(final50$modelAIC) 
table(final50$modelBIC)
table(final50$modelTIC)
```







# Part C



# Appendix